---
title: "Before-After and BACI-Analysis for AB Environment Restoration Project"
author: "Carl James Schwarz"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
    fig_caption: true
  word_document:
    reference_docx: SampleRmarkdown-Template-STYLE.docx
    toc: yes
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(width=200)

library(emmeans)
library(ggforce)
library(ggplot2)
library(ggstance)
library(lmerTest)
library(plyr)

# Load the dianostic plots for linear mixed models
# The most up to date version is located at
#source("http://www.stat.sfu.ca/~cschwarz/Stat-650/Notes/MyPrograms/schwarz.functions.r")
# but a copy is included in this directory.
source("schwarz.functions.r")

# define a simple formatter for p.values
# see https://stackoverflow.com/questions/8442322/the-use-of-format-pval-in-r-and-with-sexpr-in-sweave
format.p <- function(p, precision=0.001) {
  digits <- -log(precision, base=10)
  p <- formatC(p, format='f', digits=digits)
  p[p == formatC(0, format='f', digits=digits)] <- paste0('<', precision)
  p
}


set.seed(2342334)

n.sims.post <- 1000 # size of posterior sample

```
# Introduction
Extensive off-highway vehicle (OHV) activities can degrade fish habitat.
Reclamation activities and reductions in public motorized access 
have been initiated in Alberta to support native trout recovery. 
In Alberta, three watersheds are targeted for remediation: 
Rocky Creek, Fall Creek, and Mackenzie Creek. 
The Alberta Ministry of Environment and Parks is monitoring fish populations in these watersheds 
to examine if the reclamation activities result in a 
measurable change to the density of immature and adult fish, particularly bull trout.
The density of fish will be measured using electro-fishing methods as
described in AEP (2018) and the catch per unit effort (CUE) 
using a standardized protocol will be the response variable.

The purpose of this document is to demonstrate how the collected data will be analyzed to
examine if change has occurred in the mean CUE. Because restoration activities are ongoing,
there is little/no data currently collected post-restoration. Consequently, simulated data will be used
to demonstrate the analyses.

# Model fitting

## Sources of variation
There are several levels of variation in the models in this project.

* Stream-to-stream variation. Some streams are naturally more productive than other
streams. In any model that combines results from several streams (e.g. the BACI models), 
a random effect for stream should be present in the statistical model. 
Fortunately, because the same stream is repeatedly measured over time, the stream
serves as its own control and so the stream-to-stream variation has little impact on the
precision of the results.

* Year-to-year variation. CUE can vary across years at the same location due to 
year-specific effects (also known as process error) that affect all streams simultaneously.
For example, a colder-than-normal spring may depress water temperatures during the
sampling window which could have unspecified effects on the CUE. 
These year-specific effects tend to push all of the data for a year upwards
or downwards, which violates a key assumption of independence in regular regression
and ANOVA models as outlined in Schwarz (2018, 2017). The presence of 
year-specific effects is accounted for through a random effect of year in the models below.

* Stream-year interaction variation. Ideally, while the mean level of the
response in different stream would respond in parallel to year-specific effects,
this is unlikely to occur. The stream-year interaction variation measures
the non-parallelism in response of streams to year-specific effects. This variation
is often the limiting factor in the power of a design to detect effects -- refer to Schwarz (2018, 2017)
for more details.

* Site-to-Site variation within a stream. The CUE on a stream in a specific year will not be the
same at each site measured. This lowest level of variation represents site-specific effects
for a particular year, sampling artefacts, measuring error, etc. Multiple sites measured on a
stream in a year are pseudo-replicates. Consequently, the values should either be averaged before
the analysis, or a mixed model that incorporates stream-year random effects need to be fit. If the
designs below are balanced, e.g. every stream measured at the same number of sites in each year, the
two approaches are mathematically equivalent.


## Mixed-linear models
To account for these sources of variation, mixed linear models will be used with appropriate random
effects as needed. See the individual analyses below for more details.

The $\mathit{lmerTest}$ package (Kuznetsova et al., 2018) in the $R$ software system (R Core Team, 2018) has been used.
This package uses Restricted Maximum Likelihood (REML) to fit the models and computes the usual
estimates, standard errors, and p-values for effects. Post-hoc multiple comparisons among
effects are computed using the $\mathit{emmeans}$ package (Lenth, 2019) which provides a unified way
to estimate, for example, pair-wise comparisons such as difference in the mean between the before- and after-
restoration periods, or BACI effects without relying on 
knowledge of $R$'s internal coding system.

## Checking assumptions
It is always important to do a model assessment to ensure that the assumptions of the model are satisfied.
In this analysis the key assumptions are

* Each of the random effects (residual-effects, year-effects and site-effects) have a normal distribution
* There is no autocorrelation over time

It is possible to extract the estimates of the residuals and random effects and create
residual and caterpillar plots. A custom plotting function (*sf.autoplot.lmer*), suitable for mixed linear models and 
patterned after similar plots in the
*ggfortify* package for linear models, is available at:
[http://www.stat.sfu.ca/~cschwarz/Stat-650/Notes/MyPrograms/schwarz.functions.r]
and are utilized within this report.  A copy of the function is included in the directory with this report.

```{r echo=FALSE,message=FALSE,warning=FALSE}
# generate six scenarios; 2 sample sizes x no violations, a single outier, non-normality
# Before/after study with 4 years before and after
n.years<-8
sd.year <- .5
sd.site <- .3

mlm.data <- expand.grid(n.sites.per.year=c(4,10), Year=1:n.years)
mlm.data <- plyr::adply(mlm.data, 1, function(x) {
    x <- x[ rep(1,x$n.sites.per.year),]
    x$Site <- paste("S",'.',x$Year[1],'.',1:nrow(x), sep="")
    x})
year.eff <- data.frame(Year=1:n.years, year.eff=rnorm(n.years, sd=sd.year))
mlm.data <- merge(mlm.data, year.eff, all.x=TRUE)  # add  year effects

mlm.data$BA <- "Before"
mlm.data$BA[ mlm.data$Year > n.years/2] <- "After"
#xtabs(~BA+Year, data=mlm.data)
BA.eff <- data.frame(BA=c("Before","After"), ba.eff=c(0, .7))
mlm.data <- merge(mlm.data, BA.eff, all.x=TRUE)

mlm.data$logCUE <- 2 + mlm.data$ba.eff + mlm.data$year.eff 

# Generate the various data sets
# No violations of assumptions for residuals
mlm.data1 <- mlm.data
mlm.data1$logCUE <- mlm.data1$logCUE + rnorm(nrow(mlm.data1), sd=sd.site)
mlm.data1$Assumptions <- "No violations"

# A single outlier for a residual
mlm.data2 <- mlm.data
mlm.data2$logCUE <- mlm.data2$logCUE + rnorm(nrow(mlm.data2), sd=sd.site)
mlm.data2$logCUE[ mlm.data2$Year==1 & mlm.data2$Site=="S.1.1"] <- 
  mlm.data2$logCUE[ mlm.data2$Year==1 & mlm.data2$Site=="S.1.1"] +8
mlm.data2$Assumptions <- 'Outlier'

# A non-normal distribution e.g. t with 2 df
mlm.data3 <- mlm.data
mlm.data3$logCUE <- mlm.data3$logCUE + rt(nrow(mlm.data3), df=3)*sd.site
mlm.data3$Assumptions <- 'Non normal site-effects'

mlm.all <- plyr::rbind.fill(mlm.data1, mlm.data2, mlm.data3)
#head(mlm.all)

# Fit the before/after model to the simulated 
ba.sim.fits <- plyr::dlply(mlm.all, c("Assumptions","n.sites.per.year"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(logCUE ~ BA + (1|YearF), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA)
   BA.diff.log <- summary(pairs(fit.emmo), infer=TRUE)
   BA.ratio    <- summary(pairs(fit.emmo, type="response")) 
   list(fit=fit, 
        Assumptions=x$Assumptions[1], n.sites.per.year=x$n.sites.per.year[1],
        BA.diff.log=BA.diff.log, 
        BA.ratio=BA.ratio)
})
```

### A typical diagnostic plot
A typical diagnostic plot for one of the models (a before/after study) is:
```{r message=FALSE,warning=FALSE}
plot(sf.autoplot.lmer(ba.sim.fits[[1]]$fit))
```

The two plots in the top row examine the distribution of the residuals, the difference between the observed and fitted data.
The upper left plot should show a random distribution around 0. An outlier will show itself 
as point(s) that are far from the rest of the points. The upper right plot is a normal
probability plot. If the residuals follow a normal distribution, the points should lie close to the
reference line. Non-normality and outliers will show themselves as point(s) far 
from the line, or a systematic departure from the reference line, especially in the tails.

Outliers should be identified because they have the potential to influence the estimates when 
the number of sites per year is small. When the number of sites in year becomes larger, the
impact of an outlier is lessened.

The impact of non-normality is substantially less. In most cases, 
year-specific effects (process error) are the limiting factor that determines the
precision of the estimates and the power to detect effects. Furthermore, because sites in a 
stream-year are pseudo-replicates, the analysis essentially uses the average logCUE
over sites in a stream-year in the analysis. The Central Limit Theorem indicates that averages
of any distribution tend towards normality as sample sizes increase, so the impacts of non-normality
in the residuals
are expected to be slight.

The remainder of the plots are *caterpillar plots*, one for each random effect (aside from the residuals) in 
the model. The blue dots are the estimated value of the specific value for a level of the random effect 
(in this case for a particular year) and the whiskers represent the uncertainty in this value. If the
random effect variation is small, the blue dots should be almost vertical along the 0 (vertical) line.

The blue dots should show an approximate normal distribution around the 0 line (you need to 
imagine the density of the blue dots projected downwards). However, unless you have 20+ levels
of the random effect, you are unlikely to be able to detect non-normality.

A single random effect value that stands out
substantially from the rest may be a bit worrisome -- it represents a level of the random factor
(in this case a year) where the year-specific effect is an "outlier". Outliers can
lead to larger standard errors in the estimates and reduced power to detect effects in before/after
studies because
it says that the mean logCUE for a stream next year could be substantially different than the mean logCUE this year
due to factors out of your control. Fortunately, in BACI designs, the impact of a large year-specific
effect occurs on both the impact and control streams and so "cancels" out when the BACI comparison is done. 

The question arises as to whether such outlier random effects should be removed, in this case, by dropping a year of data? There
is no simple answer to this -- but if the year-specific effect is caused by a 1-in-a-thousand-year-flood,
there are good reasons to drop this year. 

Similarly, random effects for streams (not shown) would be interpreted in the same way. Again, because
all streams are measured before and after the restoration action, stream-effects again "cancel" in the BACI analysis 
(each stream serves as its own block) and so "outliers" in the caterpillar plots for stream-effects are
of no consequence.

Finally, random effects for year-stream interactions in the BACI model (not shown) represent
non-parallelism over time in the response of a stream to year-specific effects. Here it is important
to scrutinize outliers with care because the year-stream interaction variation drives
the precision and power of the design. 

In many cases, the simplest approach is to run the analysis with the data in question
included and excluded and see if the final conclusion changes.



### Impacts of violations at the residual level.
We simulated six datasets in a before/after scenario -- similar findings are expected for the other models in
this documents. There were `r n.years` years in the study, with half in the *before* restoration period and half in the *after*
restoration period. In each year, we simulated either 
`r sort(unique(mlm.all$n.sites.per.year))[1]` or
`r sort(unique(mlm.all$n.sites.per.year))[2]` sites per year.
Three scenarios were considered - no violations of assumptions; a single (large) outlier; and a non-normal
distribution for the residuals that is symmetric but has a much wider spread then a normal distribution.

Plots of the simulated data are:

```{r echo=FALSE}
mlm.all$n.sites.per.year.text <- paste("Sites per stream-year ", mlm.all$n.sites.per.year)
ggplot(data=mlm.all, aes(x=Year, y=logCUE))+
   ggtitle("Simulated data for residual violations in a BA experiment")+
   geom_point( position=position_jitter(width=.1))+
   facet_grid(Assumptions~n.sites.per.year.text)+
   geom_vline(xintercept=n.years/2+0.5,linetype=2)
```

The outlier in the first year is evident in the two lower plots; the impact of non-normality is not obvious in the middle two plots
and could look like (small) outliers as seen in the plots. With only a small number of sites measured in a stream-year,
it is unlikely that you will detect non-normality based on a plot of the data.

We fit the before/after model (see below) to the simulated data and extract the residual plots
(the two top most plots in the diagnostic plot from the previous section).

```{r echo=FALSE,message=FALSE,warning=FALSE}
# Extract the residual diagnostic plots
ba.sim.resid.plot <- plyr::l_ply(ba.sim.fits, function(x){
   diag.plot <- sf.autoplot.lmer(x$fit, which=c(1,2))
   diag.plot <- arrangeGrob(top=paste(x$Assumptions,"; ", x$n.sites.per.year,' sites per year',sep=""), diag.plot)
   plot(diag.plot)
})

```

The presence of the outlier is evident in either of the plots of residuals vs. fitted values or 
the normal probability plots regardless of sample size. The non-normality shows
itself only in the normal probability plots and is more evident with 10 sites per year.

The estimated effect sizes were computed for each simulated dataset:

```{r echo=FALSE}
# Get all of the fits and store them in a list with the emmeans computatons
ba.sim.diffs <- plyr::ldply(ba.sim.fits, function(x){x$BA.diff.log})
#ba.sim.diffs

# pretty up the output
temp<- ba.sim.diffs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp[, c("Assumptions","n.sites.per.year","contrast","estimate","SE","df","p.value")]
```

The effect of the outlier is to cause a negative bias
in the estimate because the outlier occurred in the *before* period and so increased the estimates
of the mean in the *before* period relative to the *after* period. The bias decreases with increasing
sample size per year as the rest of the data swamps this outlier.

The impact of non-normality is slight when the sample size is larger. When the sample size per stream-year is
small, the non-normality of a single site can look like an outlier with similar effects.

### Impacts of outliers in random effects
As noted earlier, the impact of a large year-specific effect in a before/after study can be substantial, but
the impact is less in BACI designs because each stream is measured in each year and so stream- and year-random
effects cancel in the analysis.

We will again illustrate the impacts using simulated data - in this case a BACI design with 2 streams in each of the
impact and control groups and 4 sites measured per stream-year.

```{r echo=FALSE,message=FALSE,warning=FALSE}
# generate six scenarios; 2 sample sizes x no violations, a single outier, non-normality
# Before/after study with 4 years before and after
n.years<-8
n.streams<- 4
sd.year <- .5
sd.site <- .3
sd.stream <- .5
sd.year.stream <- .2

mlm.data <- expand.grid(n.sites.per.year=10, Year=1:n.years, Stream=paste("Stream.",1:n.streams,sep=""))
mlm.data <- plyr::adply(mlm.data, 1, function(x) {
    x <- x[ rep(1,x$n.sites.per.year),]
    x$Site <- paste("S",'.',x$Year[1],'.',1:nrow(x), sep="")
    x})

year.eff <- data.frame(Year=1:n.years, year.eff=rnorm(n.years, sd=sd.year))
year.eff$BA<- "Before"
year.eff$BA[ year.eff$Year > n.years/2] <- "After"
mlm.data <- merge(mlm.data, year.eff, all.x=TRUE)  # add  year effects

stream.eff <- data.frame(Stream=paste("Stream.",1:n.streams,sep=""),
                         stream.eff=rnorm(n.streams, sd=sd.stream),
                         CI=c("C","I"))
mlm.data <- merge(mlm.data,stream.eff, all.x=TRUE)

year.stream.eff <- expand.grid(Year=1:n.years, Stream=paste("Stream.",1:n.streams,sep=""))
year.stream.eff$year.stream.eff <- rnorm(nrow(year.stream.eff), sd=sd.year.stream)
mlm.data <- merge(mlm.data, year.stream.eff, all.x=TRUE)

#xtabs(~BA+Year, data=mlm.data)
#xtabs(~Stream+CI, data=mlm.data)

BA.eff <- data.frame(BA=c("Before","After"), ba.eff=c(0, .7))
mlm.data <- merge(mlm.data, BA.eff, all.x=TRUE)
CI.eff <- data.frame(CI=c("C","I"), ci.eff=c(0, .7))
mlm.data <- merge(mlm.data, CI.eff, all.x=TRUE)

mlm.data$logCUE <- 2 + mlm.data$ba.eff + mlm.data$ci.eff + mlm.data$year.eff +mlm.data$stream.eff +mlm.data$year.stream.eff                     + rnorm(nrow(mlm.data), sd=sd.site)

# Generate the various data sets
# No violations of assumptions for residuals
mlm.data1 <- mlm.data
mlm.data1$Assumptions <- "No violations"

# A year residual
mlm.data2 <- mlm.data
mlm.data2$logCUE[ mlm.data2$Year==1 ] <-  mlm.data2$logCUE[ mlm.data2$Year==1] +8
mlm.data2$Assumptions <- 'Outlier - Year Effect'

# A stream outlier
mlm.data3 <- mlm.data
mlm.data3$logCUE[ mlm.data3$Stream=="Stream.1" ] <-  mlm.data3$logCUE[ mlm.data3$Stream=="Stream.1"] +8
mlm.data3$Assumptions <- 'Outlier - Stream Effect'

# A stream outlier
mlm.data4 <- mlm.data
mlm.data4$logCUE[ mlm.data4$Stream=="Stream.1" & mlm.data$Year==1 ] <-  mlm.data4$logCUE[ mlm.data4$Stream=='Stream.1' & mlm.data$Year==1] +8
mlm.data4$Assumptions <- 'Outlier - Year-Stream Effect'


mlm.all <- plyr::rbind.fill(mlm.data1, mlm.data2, mlm.data3, mlm.data4)
#head(mlm.all)

# Fit the before/after model to the simulated 
baci.sim.fits <- plyr::dlply(mlm.all, c("Assumptions"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   fit <- lmerTest::lmer(logCUE ~ BA + CI + BA:CI + (1|YearF) + (1|Stream) + (1|Stream:YearF), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA:CI)
   BACI.diff.log <- summary(contrast(fit.emmo,list(baci=c(-1,1,1,-1))), infer=TRUE)
   list(fit=fit, 
        Assumptions=x$Assumptions[1],
        BACI.diff.log=BACI.diff.log)
})
```

A plot of the simulated data:

```{r echo=FALSE}
ggplot(data=mlm.all, aes(x=Year, y=logCUE, color=CI, linetype=Stream))+
   ggtitle("Simulated data for BACI analysis")+
   geom_point()+
   geom_line()+
   geom_vline(xintercept=.5+n.years/2,linetype=2)+
   facet_wrap(~Assumptions, ncol=2)+
   xlab("Year\nVertical line delineates before/after")
```

shows four scenarios where an outlier exists on one of the random components. 

The diagnostic plots from the four scenarios appear below:

```{r echo=FALSE,message=FALSE,warning=FALSE}
# Extract the residual diagnostic plots
baci.sim.resid.plot <- plyr::l_ply(baci.sim.fits, function(x){
   diag.plot <- sf.autoplot.lmer(x$fit, which=c(3,4,5))
   diag.plot <- arrangeGrob(top=paste(x$Assumptions,sep=""), diag.plot)
   plot(diag.plot)
})

```

The outlier is apparent in all the plots except for that of stream.
With only a few streams it is difficult to detect
a large stream-effect, but as noted previously, it does not affect the BACI analysis.

Estimates of BACI effect sizes can be computed under each scenario:

```{r echo=FALSE}
# Get all of the fits and store them in a list with the emmeans computatons
baci.sim.diffs <- plyr::ldply(baci.sim.fits, function(x){x$BACI.diff.log})
#baci.sim.diffs

# pretty up the output
temp<- baci.sim.diffs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp[, c("Assumptions","contrast","estimate","SE","df","p.value")]
```

Only the outlier in the stream-year random effects has any influence on the estimates or 
the p-values for the
reasons explained earlier.


## Same or different sites measured over time
Sampling at the lowest level, i.e., at the site, can be done using a new set of sites for
each stream-year combination (consistent with previous sampling protocols) or by revisiting
the same sites over time (proposed changes to the sampling protocols), or a mixture of the two.

The analyses presented below implicitly assume that sites change from year-to-year within a stream. If 
some sites are revisited, this approach is slightly inefficient (i.e. reported standard errors
and reported p-values will be slightly too large) because revisiting sites over time
can account for some of the site-to-site variation in measuring trends. However, in many cases, the
limiting factors to detecting trends are year-specific effects (process error) and stream-year interactions
and so the impact is expected to be slight.

Modifications to the statistical models to account for some sites being repeated measured over time would
involve the addition of a new term representing the specific sites visited (e.g. $(1|Site)$) where the $\mathit{Site}$ 
variable is an alphanumeric code for the individual sites that remains consistent
over years (e.g. based on latitude-longitude of the site location). This addition to the model is appropriate if all sites
are repeated measured over time, or if only some of the sites are repeatedly measured over time. If none
of the sites have been measured over time, then this term is confounded with the residual variation and can 
be deleted (as shown in the models below).

## Same or different years in BACI designs
The tradition BACI analysis with multiple years before and/or after the restoration occurs, assumes
that all sites are measured in each year. The years do not have to be consecutive; for example, sampling could
occur in  2015, 2017, and 2018  before restoration, and the BACI analyses presented below would be unchanged.

The BACI analysis below also allows for missing sampling in some sites in some years. For example, suppose 
restoration occurs between 2018 and 2019.
Control site 1 could be measured in 2015, 2017, 2018, 2019, 2020 and 2021; control site 2 could be measured in
2015, 2018, 2019, 2020; and the restored sited measured in 2015, 2018, 2019, 2021 assuming the restoration occurred
between 2018 and 2019. 

However, a design where control sites are measured in 2015, 2017, 2019, and 2021 while restoration sites are
measured in 2016, 2018, 2020, and 2022 with restoration taking place between 2018 and 2019 will cause 
problems in the analysis. Technically, the design is not *connected*. In unconnected designs, there is no
contrast among the observations that leads to estimates of the BACI effect that are not confounded with year- and/or site-effects.

For example, consider the first set of missing values above where restoration takes place between 2018 and 2019 
and an X indicates that data are collected in that stream-year.

Site            |  2015  | 2016 | 2017 | 2018 | 2019 | 2020 | 2021
----------------|--------|------|------|------|------|------|-----
Control site 1  |   x    |      |  x   |  x   |   x  |  x   |   x
Control site 2  |   x    |      |      |  x   |   x  |  x   |   
Restoration site|   x    |      |      |  x   |   x  |      |   x

This design is connected because there exists at least one BACI contrast that is free of year- and site-effects.
For example, the contrast
$$(\mu_{cs1,2015}-\mu_{cs1,2018})-(\mu_{rs,2015}-\mu_{rs,2018})$$
has an "expectation" of
$$\mu_{CB}+CS1+Y2015 -\mu_{CA}-CS1-Y2018-\mu_{IB}-RS-Y2105 + \mu_{IA}+RS+Y2018=\mu_{CB} -\mu_{CA}-\mu_{IB}-+ \mu_{IA}$$
which is the BACI contrast of interest, and where 
$\mu_{CB}$, $\mu_{CA}$, $\mu_{IB}$, $\mu_{IA}$ are the means of the control sites before
restoration, control sites after restoration, restoration sites before restoration,
and restoration site after restoration; $Y2015$ (and similar) are the year-specific effects of 2015 and other
years; and $CS1$ $CS2$, and $RS$ are the site specific effects. 

But now consider the second set of missing values above where there are no common years
in the control or restoration site and restoration takes place between 2018 and 2019. 

Site            |  2015  | 2016 | 2017 | 2018 | 2019 | 2020 | 2021
----------------|--------|------|------|------|------|------|-----
Control site 1  |   x    |      |  x   |      |   x  |      |   x
Control site 2  |   x    |      |  x   |      |   x  |      |   x
Restoration site|        |   x  |      |  x   |      |   x  |   

Now there is no contrast whose "expectation" is the BACI contrast that is
free of site- and year-effects. Because no years are measured in common
between the control and restoration sites, it is not possible to obtain contrast
that is free of year-effects; it is possible to obtain a contrast that is
free of site-effects because both control and restoration sites are measured in
at least on year before and after restoration occurs.

Note that I have not used a strict definition of "expectation" because if you are willing
to believe that years and site-effects are random effect with mean 0, even the design with
alternating year is "connected" because the actual expectation of the year-specific
effect and site-specific effect is zero. But I would hesitate to use it as the actual
BACI contrast will have a much large uncertainty because it includes additional variation 
from the random effects that "don't cancel".

In designs with missing data, all of the data are used. For example, if only one site is
measured in one year (and hence is not suitable for a BACI contrast), it still provides information
about the year-to-year standard deviation. This occurs automatically in the (restricted) maximum likelihood
estimation procedure. 

So, it is not "critical" that all sites be measured in all years as long as the
resulting design is connected. This will be true if there is at least 
one year before and one year after where the control and restoration sites are measured concurrently,
but there are other combinations which are also connected. Please consult me if you have
a design with many "holes".



## Bayesian belief
### Interpretation of posterior Bayesian belief
While traditional mixed linear models provide p-values and estimates of effect size with
measures of uncertainty (standard errors and confidence intervals), the results are often
difficult to interpret.

In the traditional Neyman-Pearson system, there either is evidence ($\textit{p-value}<.05$) or
no evidence ($\textit{p-value}>.05$) of an effect. While confidence intervals provide some information
on the size of the effect, they are also difficult to deal with in a intuitive fashion. For example,
if a confidence interval for an effect size ranges from $(-1 \rightarrow 1)$, it covers the value of 0 
(indicating no evidence of an effect) but such a result is qualitatively different from another interval
$(-0.1 \rightarrow 1.9)$ that "barely" contains the value of 0. 

A Bayesian posterior belief of an effect has a very intuitive interpretation as the weight of evidence for
a hypothesis from the data, unlike a p-value which can only be interpreted as the probability of seeing this data if there was
no effect. So a posterior belief of 0.95 indicates very strong evidence of an effect (given the prior and the data collected), while
a posterior belief of 0.30 indicates very weak evidence of an effect (given the prior and the data collected). 

As noted above the posterior belief depends both on the data and the the prior distribution for the effect size.
If you have a very strong prior belief in the effect, then it will take much data to move your posterior belief 
from the prior. If you have very little prior knowledge of the effect size, then a vague (or non-informative) prior
is used, the data tends to dominate the prior, and the posterior belief is based mostly on the data collected.



### Computing posterior distributions
The posterior distribution is a complex function of the data (the likelihood) and the prior distribution
(prior belief) and is usually difficult to obtain. However,  and Hill (2006, p. 142-143) show
that the estimates from a mixed model fit are also the modes of the posterior distribution
for that parameter with suitably chosen, usually non-informative priors. 
Gelman and Su (2019) implement these ideas
in the $\mathit{arm}$ package so that a sample from the posterior distribution can 
be generated after a mixed model fit without having to use specialized software, but at 
the price that it is difficult to modify the priors. [With enough data, the priors
will have limited impact on the results]. Then it is relatively easy to compute, for example,
the Bayesian posterior belief that the difference in means between before and after restoration is
positive which has a natural interpretation. Gellman and Hill (2006, p. 347) discuss
the advantages and disadvantages of non-informative priors. 

There are various reasons to specify priors, but the usual reason is that with very sparse
data, the actual data is not very informative about effects and it is often difficult to 
fit complex models with multiple random effects. In this case, the prior can helping to stabilize computations. 
Additionally, priors can incorporate important information into an analysis that does not enter through the data, such 
as prior knowledge from many other restoration projects about the likely magnitude and direction of the effect
of restoration or the size of the year-specific effects. 
The danger of strong priors is that the results are essentially independent of the data!
With sparse data, there is no easy way to proceed but this is a natural consequence
of having very little information in the sparse data.

The $bayesglm()$ function from the $arm$ package, or  the $stan\_lmer()$ function from the $rstanarm$ package (a successor
to the $arm$ package) provided additional flexibility where you can specify the
prior distribution for the regression coefficients as well as any unknown covariance parameters.
It seems unlikely that these will be needed in this context.


# Characteristics of the simulated data
Data are used to illustrate the analyses below using a combination of real and simulated data.
All of the current data are treated as "Before" restoration data and streams are divided into their control/impact (impact = restoration activities)
status.

The simulated data starts in 2018 and ends in 2023. The simulated 2018 year is treated as an additional "Before" year so that 
all streams have at least two years of "Before" data (this is needed to fit the broken-stick model below).
[When the actual data are available for the full experiment, the start of the restoration period can be set separately for each watershed.]

The number of sites simulated in each stream-year is a random sample taken from the set of number of sites measured
in each steam-year in the current data with a minimum of two sites measured per stream-year. The first
four (or the actual number if less than 4 sites are simulated) are treated as "permanent" sites that are 
repeatedly measured over time. All other sites are assumed to be new sites in each stream-year.

The existing data was used to estimate the variance components associated with streams, years, and stream-year interactions.
These variance components were used to generate random normal deviations on the $log(\mathit{CUE})$ representing
the effects of streams, year, and stream-years.

The current data has virtually NO sites that were repeatedly measured over time, so it is not possible to
estimate the site variance component.
Consequently, no site-effects were generated so it will not be surprising that 
models with *Site* effects will have estimated site variance components of zero. [You can still fit the models
with permanent sites -- the results would then be essentially the same as the model without site-effects.]

## Simulated effect sizes
### Additive step change
A simulated step effect of restoration was added to the simulated data above for the restoration streams but only for one species. The size of the step effect ($\mu_\textit{after}-\mu_\textit{before})$ on the $\log()$ scale is:

```{r echo=FALSE}
effect.size.mean <- read.csv("effects.size.mean.csv", header=TRUE, as.is=TRUE, strip.white=TRUE)
effect.size.mean
```

### Change in slope
The control sites were assumed to have no change in slope between the before and after periods for
all species. A slope change ($\beta_\textit{after}-\beta_\textit{before}$) was used to modify the CUE (on the log-scale) as shown below:
```{r echo=FALSE}
effect.size.slope <- read.csv("effects.size.slope.csv", header=TRUE, as.is=TRUE, strip.white=TRUE)
effect.size.slope
```

The simulated data was written to a *.csv* file for usage in this document. 




# Preliminary Data Preprocessing
Data will be available from the FWMIS system and it is assumed that 
it will have been downloaded into a suitable format, e.g.
an Excel workbook or a *csv* file. We currently use the simulated data as described above.

## Imputation of 0 values
One of the key issues with many database systems is accounting for sites that are visited, but NO fish, of a particular/all species were 
detected. In many cases, database systems only record positive counts. A naive extraction of the data from such systems would 
not capture the 0 counts for species looked for but not seen.

This could lead to biases in estimates. For example, consider the very simple example where:

Site | Species | Count
-----|---------|------
A    | RNTR    |  3
A    | BKTR    |  2
B    | BKTR    |  2
B    | BLTR    |  5
C    | RNTR    |  5
C    | BKTR    |  2
----------------------

An incorrect estimate of mean density over the
three sites for RNTR would be $(3+5)/2$ rather than the correct estimate of $(3+0+5)/3$.

To account for species with 0 counts, the database system must have two separate tables.
In the first table, is a record of all visits made to all sites in each year regardless if any fish are captured or not. 
This table should contain information
that is specific to the visit, such as the length of the segment, the personnel used, etc.

The second table would contain a count of the detected fish species 
for each visit. Notice that only records for species with positive
counts will be included. Consequently, it will be necessary to impute counts of 0 for sites in which a visit was made but
this species was not detected.

By merging the two tables, records with 0 counts for species not detected can be inserted.
This document will assume that this *zero imputation* has been done, 
so the input file has a record of the count (including 0) for each species 
of interest at each year-site-visit combination. 
The CUE is then computed as the $count / distance~sampled * 300$ to standardize per 300 m of effort for each visit.

The current FWMIS system has been set up as a relational database along these lines.
One table has higher level information about a site (distance shocked, location, etc.) 
and a second table has dependent information on species captured (fork length, weight, sex, etc.). 
The Oracle query in FWMIS is now designed to provide information on sites with zero fish encountered to account for
this potential problem.

## Additional variables needed
Each combination of stream and year also needs to be labelled with two variables. 
The variable $\mathit{BA}$ indicates if the measurement 
takes place *before* or *after* the restoration
takes place in the watershed (e.g. using codes $Before$ and $After$); the variable $\mathit{CI}$ indicates
if the stream is a control ($C$) or restored (treated, $T$). 

Of course, other codes can be used and the code below may have to be modified. 
The modifications should be obvious in most cases. Some care will be needed
in the sections of code where the Bayesian posterior beliefs are extracted as
the variable name used to extract the posterior samples depends on the internal 
$R$ coding of factor levels. A simple listing of the few records from the data frame generated
by the *arm::sim()* function will show the variable names generated by $R$.



## Response variable
As explained in previous reports, the response variable is the 
$log(\mathit{CUE}+\textit{offset})$ where $log()$ refers to the natural
logarithmic transform and the $\textit{offset}$ is 1/2 of the smallest non-zero value. 
A separate offset will be used for each
combination of species, watershed, and age-class.

## Read data
We read in the (simulated) CUE data that includes the before and after data and
compute the $log(CUE+offset)$.
```{r }
cue.data <- read.csv("simulated.data.csv",  header=TRUE, as.is=TRUE, strip.white=TRUE)
head(cue.data)
tail(cue.data)


# compute offset for each watershed, species, age-class
offset <- plyr::ddply(cue.data, c("Watershed","Species","AgeClass"), plyr::summarize,
                      offset = .5* min(CUE[ CUE>0]))
offset


cue.data <- merge(cue.data, offset)
cue.data$logCUE <- log(cue.data$CUE + cue.data$offset)
```

I'm a paranoid guy, so I do extensive checks before doing any analysis!

## How many sites measured each year on each stream?
We first enumerate how many sites are measured on each stream in each year.

```{r }
stream.year.sites <- unique(cue.data[,c("Watershed","Stream","Year","Site")] )
xtabs(~paste(Watershed,',',Stream,sep="")+Year, data=stream.year.sites)

```
The actual number that are measured for each stream-year is not that 
important and the analyses below do not depend on the design being balanced,
i.e. different number of sites can be measured for each stream-year 
,and in some years, no sites could be measured.

## Check BA classification
We check that the $\mathit{BA}$ variable has been set for each stream-year combination. 
First, any one year must be classified
into one and only one period.
```{r }
stream.year <- unique(cue.data[,c("Watershed","Stream","Year","BA")] )
xtabs(~paste(Watershed,',',Stream,sep="")+Year, data=stream.year)
```
We check that the above table has only 1 and 0's for any entry. Any entry that is >1 indicates a year that is classified
in more than one period for  particular stream, which is not possible.

Similarly, we look at the actual classification of each stream-year combination:
```{r }
# R allows matrices to be index by the row/column names. We set up a matrix with the appropriate names
# and then assign the values based on the combination of row names and column names seen in the data.
# See the help files for '[' for more information
stream.year <- unique(cue.data[,c("Watershed","Stream","Year","BA")] )
BA.table <- matrix(" ", nrow=length(unique(cue.data$Stream)), ncol=length(unique(cue.data$Year)))
colnames(BA.table) <- as.character(sort(unique(cue.data$Year)))
rownames(BA.table) <- sort(unique(paste(cue.data$Watershed,'.',cue.data$Stream,sep="")))
BA.table[ cbind( paste(stream.year$Watershed,".",stream.year$Stream,sep=""), as.character(stream.year$Year))] <- substr(stream.year$BA,1,1)
BA.table
```
We check that for all years, the correct designation of before and after has been done.

## Check CI classification
Similarly, we check that the $\mathit{CI}$ variable has been set for each stream-year combination. First, any one stream-year must be classified
into either the treatment or control groups.
```{r }
stream.year <- unique(cue.data[,c("Watershed","Stream","Year","CI")] )
xtabs(~paste(Watershed,',',Stream,sep="")+Year, data=stream.year)
```
We check that the above table has only 1 and 0's for any entry. Any entry that is >1 indicates a year that is classified
in more than one of treatment or control for  particular stream, which is not possible.

Similarly, we look at the actual classification of each stream-year combination:
```{r }
stream.year <- unique(cue.data[,c("Watershed","Stream","Year","CI")] )
CI.table <- matrix(" ", nrow=length(unique(cue.data$Stream)), ncol=length(unique(cue.data$Year)))
colnames(CI.table) <- as.character(sort(unique(cue.data$Year)))
rownames(CI.table) <- sort(unique(paste(cue.data$Watershed,'.',cue.data$Stream,sep="")))
CI.table[ cbind( paste(stream.year$Watershed,".",stream.year$Stream,sep=""), as.character(stream.year$Year))] <- substr(stream.year$CI,1,1)
CI.table
```
We check that for all years, the correct designation of *control* or *impact* has been done.

## Check for sites measured more than once.
We check for repeated sites.
```{r }
# Check for repeatedly measured sites over time
sites <- unique(cue.data[,c("Watershed","Stream","Site","Year")])
n.visits.per.site <- plyr::ddply(sites, c("Watershed","Stream","Site"), plyr::summarize,
                                 n.visits=length(Year))

cat("Sites with more than one visit over time\n")
rep.sites <- n.visits.per.site[ n.visits.per.site$n.visits >1,]

head(rep.sites)
sites.red <- merge(sites, rep.sites, all.y=TRUE) # only selected the repeated sites.

xtabs(~ paste(Watershed,'.',Stream,'.',Site,sep="")+Year, data=sites.red, exclude=NULL, na.action=na.pass)
```
We see that there were essentially no repeated sites in the real data (2017 and before),
but up to 4 repeated sites per stream in the simulated data (sites labelled SP1, SP2, SP3, or SP4).
Not all repeated sites are measured in each year.


```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```
# Before-After Analysis
The simplest analysis is a before-after analysis where the mean CUE is compared between the before restoration period
and the after restoration period. This assumes that restoration results in a step change in restoration, rather than a gradual trend.
No control river is used, so an effect may be confounded with temporal changes between the two periods.

## Preliminary plot
A preliminary plot is constructed to check for outliers etc. 

```{r }
# Compute the mean logCUE for each year
mean.CUE <- plyr::ddply(cue.data, c("Watershed","Stream","Species","AgeClass","Year","BA"), 
                plyr::summarize,
                    mean.logCUE=mean(logCUE, na.rm=TRUE))

# When did restoration take place for each watershed/species
year.restore <- plyr::ddply(mean.CUE, c("Watershed","Stream","Species","AgeClass"), plyr::summarize,
                            year.restore = max( Year[ BA == "Before"]))

# A separate plot is produced for each species
plyr::l_ply(sort(unique(cue.data$Species)), function(x,cue.data, mean.CUE){
  sel.cue.data <- cue.data[ cue.data$Species==x,]
  sel.mean.CUE <- mean.CUE[ mean.CUE$Species==x,]
  plot.cue<- ggplot(data=sel.cue.data, aes(x=Year, y=logCUE, color=AgeClass))+
     ggtitle(paste("Preliminary plot of the data for ",x))+
     geom_point( position=position_jitter(h=.1))+
     geom_line(data=sel.mean.CUE, aes(y=mean.logCUE), position=position_jitter(h=.1))+
     facet_wrap(~paste(Watershed, '.', Stream), ncol=3)+
     theme(legend.justification=c(1,0), legend.position=c(1,0))
  plot(plot.cue)
}, cue.data=cue.data, mean.CUE=mean.CUE)   
```

If outliers are detected, they should be dealt with.


## The statistical model.
The statistical model must account for the multiple measurements in a site-year (the multiple sites). This can be done
in two ways

* Average the $log(CUE+\textit{offset})$ for each site-year combination and analyze the averages using
a t-test (or similar analysis). 
This will be approximate if the design is unbalanced, but the difference from the exact
analysis (below) should, in most cases, be slight.
* Use a mixed-linear model with random year-effects to account for the multiple measurements
in a year. 

If the design were exactly balanced, i.e. the same number of measurements (sites) in each year, 
the two approaches would give exactly the same results.

We will use the mixed-linear model approach for every combination of species-ageclass-stream.
The model is
$$log(\mathit{CUE}+\textit{offset}) \sim \mathit{BA} + \mathit{YearF}(R)$$
where $\mathit{BA}$ is the before/after classification for each year, and 
$\mathit{YearF}$ is a random effect of years
(it must be declared as a factor to avoid being 
confused with a linear trend).
A multiple comparison procedure will then estimate the size of the effect on both the log-scale 
(difference in mean logarithms) and the original scale (ratio of means) between the before and after periods.

## Using R to fit and extract results
The $\mathit{lmerTest}$ package is used to fit the model; the $\mathit{emmeans}$ package is used to estimate the
period effect after the model is fit.

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
ba.fits <- plyr::dlply(cue.data, c("Watershed","Stream","Species","AgeClass"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ BA + (1|YearF), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA)
   BA.diff.log <- summary(pairs(fit.emmo), infer=TRUE)
   BA.ratio    <- summary(pairs(fit.emmo, type="response")) 
   list(fit=fit, 
        BA.diff.log=BA.diff.log, 
        BA.ratio=BA.ratio)
})

# Show the results for the first fit
ba.fits[[1]]$fit
ba.fits[[1]]$BA.diff.log
ba.fits[[1]]$BA.ratio
```
A message about a singular fit indicates that the estimated standard deviation of the year random effect for some of the streams is 0.
The estimated difference/ratio between the after and before periods, along with standard errors, a 95% confidence limit,
and p-value of the test for no effect is produced.

The fitted values and observed data can be plotted together. The blue line below is the fitted before/after effect
and the red line joins the mean of the log(CUE+offset).
```{r  }

# Get the fitted values
ba.pred <- plyr::ldply(ba.fits, function(x){
   #browser()
   # need to convert the YearF from factor back to numeric value 
   pred <- cbind( x$fit@frame, pred=predict(x$fit, re.form=~0))
   pred$Year <- as.numeric(as.character(pred$YearF))
   pred
})

# Now to generate the observed and fitted data for each Watershed/Stream combination
npages=nrow(unique(cue.data[,c("Watershed","Stream","Species")]))*3/9
plyr::l_ply(1:npages, function(page){
  plot.fit <- ggplot(data=cue.data, aes(x=Year, y=logCUE))+
     ggtitle("Observed and fitted data from before/after model")+
     geom_point(position=position_jitter(width=0.1))+
     geom_line(data=ba.pred,  aes(y=pred), color='blue', size=2)+
     geom_line(data=mean.CUE, aes(y=mean.logCUE), color="red")+
     geom_vline(data=year.restore, aes(xintercept=year.restore+.5))+
     facet_grid_paginate(Watershed+Stream+Species~AgeClass, page=page, nrow=3, ncol=3)
  plot(plot.fit)
})
```



This set of fits and estimated differences/ratio can be used to construct summary plots

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
ba.diffs <- plyr::ldply(ba.fits, function(x){x$BA.diff.log})
#head(ba.diffs)

# pretty up the output
temp<- ba.diffs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp

ggplot(data=ba.diffs, aes(y=paste(Watershed,".",Stream,sep=" "), x=estimate))+
   ggtitle("Estimated difference (log scale) between before and after")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=lower.CL, xmax=upper.CL), height=.01)+
   xlab("Estimated difference (log-scale) and 95% ci")+
   ylab("Watershed - Stream")+
   geom_vline(xintercept=0)+
   geom_text( label=format.p(ba.diffs$p.value,precision=.001), aes(x=-Inf), hjust=-.01, vjust=.5, size=2)+
   facet_grid(Species~AgeClass)


```

A similar plot can be made for the ratio of the mean (not shown). 

Case where the 95% confidence interval does not include the value of 0 (vertical line) indicate
evidence of a difference. The actual p-value is presented in the *ba.diffs* data.frame constructed 
above if these are of interest.

## Bayesian posterior beliefs
The $\mathit{arm}$ package can be used to generate samples from the posterior distribution of the effect size.
Some care is needed because knowledge of the coding used by $R$ is required -- in particular, 
is the estimate of the fixed effect the difference of means computed as 
$\mathit{mean}_\mathit{before} - \mathit{mean}_\mathit{after}$ or vice-versa.

```{r  }
# sample from the posterior, plot the violin plots and compute the posterior belief that after-before >0
ba.diffs.post <- plyr::ldply(ba.fits, function(x){
   #browser()
   post.samples <- arm::sim(x$fit, n.sims=n.sims.post)
   data.frame(post.diff.A.minus.B = -post.samples@fixef[,2]) # notice that R computes before-after
})
head(ba.diffs.post)

# compute the posterior belief that slope > 0
ba.diffs.belief <- plyr::ddply(ba.diffs.post, c("Watershed","Stream","Species","AgeClass"),
                     plyr::summarize,
                       post.belief.gt.0 = mean(post.diff.A.minus.B >0))
head(ba.diffs.belief)

ggplot(data=ba.diffs.post, aes(y=paste(Watershed,".",Stream,sep=" "), x=post.diff.A.minus.B))+
   ggtitle("Posterior sample for difference (log scale) in After - Before mean")+
   ggstance::geom_violinh()+
   geom_text(data=ba.diffs.belief, label=formatC(ba.diffs.belief$post.belief.gt.0,2, format="f"), 
             aes(x=-Inf), hjust=-.2) +
   xlab("Difference (log-scale) (After-Before)\nWith posterior belief that difference > 0")+
   ylab("Watershed - Stream")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)
```

The posterior belief that the difference in means (log-scale; After-Before) is positive has a nice
interpretation. 

## Checking assumptions
It is always important to do a model assessment to ensure that the assumptions of the model are satisfied.
In this analysis the key assumptions are

* Each of the random effects (year-effects and site-effects) have a normal distribution
* There is no autocorrelation over time

It is possible to extract the estimates of the two sets of random effects and create
residual and caterpillar plots.

```{r message=FALSE,warning=FALSE}
# The dianostic plot is produced for the first fit as a demonstration
plyr::l_ply(ba.fits[2], function(x){
   diag.plot <- sf.autoplot.lmer(x$fit)
   plot(diag.plot)
  
})



```

The residual plot (upper left corner) should show random scatter around 0. There are only two columns
to the residuals because there are only two fitted values, the mean before restoration took place
and the mean after restoration took place.

The normal
probability plot (upper right) for the residuals (the site-effects) should fall
close to the theoretical line of fit. It is not crucial that these lowest random effects have an exact normal distribution
since they are effectively averaged in the analysis and the central-limit theorem 
indicates that the average of any distribution will tend towards normality as the
sample size increases.

The caterpillar plot (bottom left) should
show no large outlier effects for specific years. In this case, the estimated standard
deviation of the year-effects is 0, so all of the points (with 95% credible interval)
are the same at 0.

It will be extremely difficult to detect autocorrelation over time unless you have 10+ years of 
data (Bence 1995) and so this is not done here.


## Adjusting the analysis for repeatedly measuring the same site over time
As noted earlier, the presented analyses assume that a different set of sites
is measured in each stream-year combination. However, all else being equal,
there is some advantage to measuring the same site repeated over time.
In this case, the above model must be modified to account for the 
site-specific effects. We assume that the $Site$ variable has a unique alphanumeric
code representing a particular site. The model is adjusted by including another
term $(1|\mathit{Site})$ that accounts for the same site being repeatedly measured over time.
Not all sites need to be repeatedly measured, and a site does not have to be repeatedly measured in all years. 
However, at least one site should have been measured in more than one year.


We then modify the previous fitting code to account for some with sites with multiple measurements
by adding the *(1|Site)* term to the model. The rest of the code can then be run without any other
changes.
```{r}
# Get all of the fits and store them in a list with the emmeans computatons
# Account for repeated measurements on the same stream
ba.fits.rep <- plyr::dlply(cue.data, c("Watershed","Stream","Species","AgeClass"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ BA + (1|YearF) +(1|Site), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA)
   BA.diff.log <- summary(pairs(fit.emmo), infer=TRUE)
   BA.ratio    <- summary(pairs(fit.emmo, type="response")) 
   list(fit=fit, 
        BA.diff.log=BA.diff.log, 
        BA.ratio=BA.ratio)
})

# Extract one of the fitting results for closer examination
# In particular, look at the variance component associated with permanent Sites
ba.fits.rep[[1]]$fit
ba.fits.rep[[1]]$BA.diff.log
ba.fits.rep[[1]]$BA.ratio

# the rest of the code has no other modifications
# Get all of the fits and store them in a list with the emmeans computatons
ba.diffs.rep <- plyr::ldply(ba.fits.rep, function(x){x$BA.diff.log})
#head(ba.diffs)

# pretty up the output
temp<- ba.diffs.rep
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp

ggplot(data=ba.diffs.rep, aes(y=paste(Watershed,".",Stream,sep=" "), x=estimate))+
   ggtitle("Estimated difference (log scale) between before and after",
           subtitle="Repeated measurements on the same site accounted for")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=lower.CL, xmax=upper.CL), height=.01)+
   xlab("Estimated difference (log-scale) and 95% ci")+
   ylab("Watershed - Stream")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)+
   geom_text( label=format.p(ba.diffs.rep$p.value,precision=.001), aes(x=-Inf), hjust=-.01, vjust=.5, size=2)



```





```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```
# Trend with a breakpoint
## Introduction
The before/after analysis assumes a step-change in the mean between the before and after
restoration activities. In some cases, the effect of restoration can be more gradual
with incremental improvements over time.

In this section, we will fit a *broken-stick* trend where the slopes of the trends
could be different between the two periods, but the lines are joined together
at the point of restoration. For example, here is sample plot illustrating the line we wish to fit:
```{r echo=FALSE}
# Generate a sample plot of broken stick line
sample.data <- expand.grid(Year=2010:2020, Site=1:5)
sample.data$logCUE <- 1 + .1*sample.data$Year + .15*pmax(0,sample.data$Year-2015) + rnorm(nrow(sample.data), sd=.2)
sample.data$Yearp <- pmax(0, sample.data$Year-2015)
# add year specific effects
sample.year.eff <- data.frame(Year=2010:2020, year.eff=rnorm(length(2010:2020), sd=.1))
sample.data <- merge(sample.data, sample.year.eff, all.x=TRUE)
sample.data$logCUE <- sample.data$logCUE + sample.data$year.eff
sample.data$YearF <- factor(sample.data$Year)
sample.data.fit <- lmerTest::lmer(logCUE ~ Year + Yearp + (1|YearF), data=sample.data)
sample.data$predict <- predict(sample.data.fit, re.form=~0)

ggplot(data=sample.data, aes(x=Year, y=logCUE))+
   ggtitle("Illustration of broken-stick line",
           subtitle="Includes year-specific effects (process error)")+
   geom_point()+
   geom_line(aes(y=predict))+
   geom_vline(xintercept=2015.1)+
   scale_x_continuous(breaks=2000:3000)+
   xlab("Year \n Vertical line is when restoration takes place")

```

Interest lies if the **slope** changes between the pre- and post-restoration activity.

## The statistical model
The statistical model must include a mechanism to create the broken-stick and to account
for year-specific effect (process error). The model is:
$$log(\mathit{CUE}+\textit{offset}) \sim Year + YearP + \mathit{YearF}(R)$$
where $\mathit{Year}$ represents the yearly trend before restoration takes place;
$\mathit{YearP}$ represents the change in trend after restoration takes place; and
$\mathit{YearF}$ again represents the (random) year-specific effects (process error).

## Using R to fit and extract results

The data structure includes a new variable $\mathit{YearP}$ which is 0 for years prior 
to when restoration takes place, and then takes the value of $\mathit{Year}-\mathit{Year~of~restoration}$.
For example, if restoration took place in 2015, the new $\mathit{YearP}$ variable is computed (in $R$) as
`data$YearP = pmax(0,data$Year - 2015)`.

The $\mathit{lmerTest}$ package is used to fit the model; the $\mathit{emmeans}$ package is used to estimate the
period effect after the model is fit.

```{r  warning=FALSE}
# Get all of the fits and store them in a list with the emmeans computatons

bs.fits <- plyr::dlply(cue.data, c("Watershed","Stream","Species","AgeClass"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   
   # find the year of restoration
   restore.year <- max(x$Year[ substr(x$BA,1,1)=="B"])
   # compute the YearP variable
   x$YearP <- pmax(0, x$Year - restore.year)
   
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ Year + YearP + (1|YearF), data=x)
   
   # extract the estimates of the two slopes and the change in the slope
   # also get the confidence intervas
   ci <- confint(fit, "Year", quiet=TRUE, method="Wald")
   init.slope <- cbind(summary(fit)$coefficients[ row.names(summary(fit)$coefficients)=="Year",,drop=FALSE], ci)
   #browser()
   ci <- confint(fit, "YearP", quiet=TRUE, method="Wald")
   delta.slope<- cbind(summary(fit)$coefficients[ row.names(summary(fit)$coefficients)=="YearP",,drop=FALSE], ci)
   # compute the final slope by fitting the same model except in reverse
   x$YearP2 <- pmax(0, restore.year-x$Year)
   fit2 <- lmerTest::lmer(log(CUE+offset) ~ Year + YearP2 + (1|YearF), data=x)
   ci <- confint(fit2, "Year", quiet=TRUE, method="Wald")
   final.slope<-cbind(summary(fit2)$coefficients[ row.names(summary(fit2)$coefficients)=="Year",,drop=FALSE], ci)
   
   # fix the names for each 
   colnames(init.slope) <- make.names(colnames(init.slope))
   colnames(delta.slope)<- make.names(colnames(delta.slope))
   colnames(final.slope)<- make.names(colnames(final.slope))
   list(Watershed=x$Watershed[1], Stream=x$Stream[1], Species=x$Species[1], AgeClass=x$AgeClass[1],
        fit=fit, 
        init.slope=init.slope,
        delta.slope=delta.slope,
        final.slope=final.slope)
})

# Show the results for one of the fits
bs.fits[[1]]$fit
bs.fits[[1]]$init.slope
bs.fits[[1]]$delta.slope
bs.fits[[1]]$final.slope
```
A message about a singular fit indicates that the estimated standard deviation of the year random effect is 0.
The estimated change in the slope between the after and before periods, along with standard errors, a 95% confidence limit,
and p-value of the test for no change in the slope is produced.


We can examine the plots of fitted and predicted values to check the general form of the fit and
again for any obvious problems:
```{r}
# Get the fitted values
bs.pred <- plyr::ldply(bs.fits, function(x){
   #browser()
   # need to convert the YearF from factor back to numeric value 
   pred <- cbind( x$fit@frame, pred=predict(x$fit, re.form=~0))
   pred$Year <- as.numeric(as.character(pred$YearF))
   pred
})

# Now to generate the observed and fitted data for each Watershed/Stream combination
npages=nrow(unique(cue.data[,c("Watershed","Stream","Species")]))*3/9
plyr::l_ply(1:npages, function(page){
  plot.fit <- ggplot(data=cue.data, aes(x=Year, y=logCUE))+
     ggtitle("Observed and fitted data from broken stick model")+
     geom_point(position=position_jitter(width=0.1))+
     geom_line(data=bs.pred,  aes(y=pred), color='blue', size=2)+
     geom_line(data=mean.CUE, aes(y=mean.logCUE), color="red")+
     geom_vline(data=year.restore, aes(xintercept=year.restore+.5))+
     facet_grid_paginate(Watershed+Stream+Species~AgeClass, page=page, nrow=3, ncol=3)
  plot(plot.fit)
})
```


This set of fits and estimated differences/ratio can be used to construct summary plots

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
bs.diffs <- plyr::ldply(bs.fits, function(x){x$delta.slope})
#head(bs.diffs)

# pretty up the output
temp<- bs.diffs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.value <- NULL
temp

ggplot(data=bs.diffs, aes(y=paste(Watershed,".",Stream,sep=" "), x=Estimate))+
   ggtitle("Estimated change in slope (log scale) between before and after")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=X2.5.., xmax=X97.5..), height=.01)+
   xlab("Estimated change in slope (log-scale) and 95% ci")+
   ylab("Watershed - Stream")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)+
   geom_text( label=format.p(bs.diffs$Pr...t.., precision=.001), aes(x=-Inf), hjust=-.01, vjust=.5, size=2)



```


Case where the 95% confidence interval does not include the value of 0 (vertical line) indicate
evidence of a difference in the slope. The actual p-value is presented in the *bs.diffs* data.frame constructed 
above if these are of interest.

## Bayesian posterior beliefs
The $\mathit{arm}$ package can be used to generate samples from the posterior distribution of the effect size.
Here we can access the estimated difference in the slopes directly.

```{r  message=FALSE,warning=FALSE}
# sample from the posterior, plot the violin plots and compute the posterior belief that after-before >0
bs.diffs.post <- plyr::ldply(bs.fits, function(x){
   #browser()
   #cat(x$Watershed, x$Stream, x$Species, x$AgeClass,"\n") # use this to follow progress of the sampling
   #if(x$Stream=="Thistle")browser()
   post.samples <- arm::sim(x$fit, n.sims=n.sims.post)
   data.frame(post.diff.slope = post.samples@fixef[,"YearP"]) # Year P is the difference in slope (after-before)
})
head(bs.diffs.post)

# compute the posterior belief that difference (A-B) > 0
bs.diffs.belief <- plyr::ddply(bs.diffs.post, c("Watershed","Stream","Species","AgeClass"),
                     plyr::summarize,
                       post.belief.gt.0 = mean(post.diff.slope >0))
head(bs.diffs.belief)

ggplot(data=bs.diffs.post, aes(y=paste(Watershed,".",Stream,sep=" "), x=post.diff.slope))+
   ggtitle("Posterior sample for difference (log scale) in the slope (After - Before)")+
   ggstance::geom_violinh()+
   geom_text(data=bs.diffs.belief, label=formatC(bs.diffs.belief$post.belief.gt.0,2, format="f"), 
             aes(x=-Inf), hjust=-.2) +
   xlab("Change in slope (log-scale) (After-Before)\nWith posterior belief that difference > 0")+
   ylab("Watershed - Stream")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)
```

The posterior belief that the difference in slopes (After-Before) is positive has a simple 
interpretation. 

## Checking assumptions
It is always important to do a model assessment to ensure that the assumptions of the model are satisfied.
In this analysis the key assumptions are

* Each of the random effects (year-effects and site-effects) have a normal distribution
* There is no autocorrelation over time

It is possible to extract the estimates of the two sets of random effects and create
residual and caterpillar plots.

```{r message=FALSE}
# The dianostic plot is produced for one of the fits as a demonstration
plyr::l_ply(bs.fits[2], function(x){
   diag.plot <- sf.autoplot.lmer(x$fit)
   plot(diag.plot)
  
})
```

The diagnostic plot is interpreted in the same way as previously described. There are now more fitted values
because of the broken-stick line makes a prediction for each year of the study.


```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```

# BACI analysis within each watershed on the mean
## Introduction
The problem with a before/after analysis is that the observed change could be 
confounded with temporal effects. A Before-After-Control-Impact design uses one or more
control streams to adjust for temporal effects.

In this example, the Fall stream in the Ram watershed does not have any control streams
and so will be excluded from this analysis.

The key advantage of several control streams is that the scope of inference is now expanded.
If there was only a single control stream, then one could argue that the effects observed
are peculiar to the control stream chosen. If multiple control streams are (randomly) chosen,
then one can argue that the effects are generalizable to all control streams in the watershed.

Each watershed only has one treatment stream, so it can still be argued that the results are peculiar 
to the treatment stream chosen. For the same reasons, it is also possible to have multiple 
treatment streams within a watershed.

The methods presented below will work if there are multiple control and/or multiple treatment streams
but there must be replication in at least one of the groups. If a watershed only has a single treatment
and a single control stream, the model below will still be applicable by removing the $\mathit{Stream}$ 
random effects.

The final result from a BACI analysis is the **differential change** in the mean of treatment stream relative to the mean of the control 
stream(s). For example, suppose that the mean in the control streams changes from 0 to 1 between the before and after periods and
the mean in the treatment streams changes from 3 to 6. The BACI effect is the differential change, i.e. $(6-3)=3$ vs.
$(1-0)=1$ or $(6-3)-(1-0)=2$. For an analysis on the $log(\mathit{CUE})$ the BACI effect is difference in the mean
logarithm which can be translated into ratios of ratios.


## The statistical model
As in the before/after analysis, the analysis can be done in two ways.

* Average the multiple measurements from the sites first.
* Fit a mixed linear model to the original measurements.

Again, if the design were exactly balanced, the two approaches would give exactly the same results.


The statistical model for each watershed is
$$log(\mathit{CUE}+\textit{offset}) \sim \mathit{BA} + \mathit{CI} + \mathit{BA}:\mathit{CI} + \mathit{YearF}(R) + \mathit{Stream}(R)+ \mathit{YearF}:\mathit{Stream}(R)$$
where the $\mathit{BA}$ refers to the before/after effect; 
the $\mathit{CI}$ refers to the control/impact effect;
the $\mathit{BA}:\mathit{CI}$ refers to the BACI effect (the differential change) which is the term of interest;
the $\mathit{YearF}$ refers to the year random effects (it must be declared as a factor to avoid being 
confused with a linear trend);
the $\mathit{Stream}$ term refers to the (random) differences among streams;
and the $\mathit{YearF}:\mathit{Stream}$ refers to the random effect where streams do not 
track each other in parallel over time.

The effects of stream will cancel in the analysis because each stream is measured
before and after impact, i.e., each stream serves as a blocking variable.
The effects of year will also cancel in the analysis because in each year, both control
and impact streams are measured, i.e., each year also serves as a blocking variable.
The last term represents inconsistency in the mean of a stream over time compared to other streams, i.e.,
a non-parallelism effect and is often the limiting factor in a power analysis 
in a BACI design.

The same model can be used if some streams are not measured in some years and if the design were unbalanced.


## Using R to fit and extract results
We exclude the watersheds that don't have a control and treatment stream and a replicate of either the control
or treatment streams.

```{r }
n.streams <- plyr::ddply(cue.data, "Watershed", plyr::summarize,
                         n.control  =length(unique(Stream[ CI=="C"])),
                         n.treatment=length(unique(Stream[ CI=="T"])))
n.streams

watershed.exclude <- n.streams[ n.streams$n.control==0 | n.streams$n.treatment==0,]
watershed.exclude

dim(cue.data)
cue.data.red <- cue.data[ !cue.data$Watershed %in% watershed.exclude$Watershed, ]
dim(cue.data.red)
```

The $\mathit{lmerTest}$ package is used to fit the model; the $\mathit{emmeans}$ package is used to estimate the
BACI effect after the model is fit. Some care is needed to specify the coefficients of the BACI contrast 
properly. For example, in this model, the interaction (BACI) term is specified as $\mathit{BA}:\mathit{CI}$.
The $\mathit{BA}$ variable is coded using *After* and *Before*; the $\mathit{CI}$ variable is coded using *C*
and *I*. Consequently, the BACI terms are in the order *After C*, *Before C*,*After T*, and *Before T*.

We would like the BACI contrast to be computed as the difference of difference in the means as
$$\mathit{BACI} = (\mu_{\mathit{After~T}} - \mu_{\mathit{Before~T}}) - (\mu_{\mathit{After~C}} - \mu_{\mathit{Before~C}})$$
and so the BACI coefficients are $(-1,1,1,-1)$ as specified below.

```{r  warning=FALSE}
# Get all of the fits and store them in a list with the emmeans computatons
baci.fits <- plyr::dlply(cue.data.red, c("Watershed","Species","AgeClass"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ BA + CI + BA:CI + (1|YearF) +(1|Stream) +(1|YearF:Stream), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA:CI)
   BACI.diff.log <- summary(contrast(fit.emmo,list(baci=c(-1,1,1,-1))), infer=TRUE)
   BACI.ratio    <- summary(contrast(fit.emmo,list(baci=c(-1,1,1,-1)), type="response")) 
   list(fit=fit, 
        BACI.diff.log=BACI.diff.log, 
        BACI.ratio=BACI.ratio)
})

# Show the results for the first fit
baci.fits[[1]]$fit
baci.fits[[1]]$BACI.diff.log
baci.fits[[1]]$BACI.ratio
```
A message about a singular fit indicates that the estimated standard deviation of some of the random effects is 0.
The estimated BACI effect along with standard errors, a 95% confidence limit,
and p-value of the test for no effect is produced.

We can examine the plots of fitted and predicted values to check the general form of the fit and
again for any obvious problems:
```{r}
# Get the fitted values
baci.pred <- plyr::ldply(baci.fits, function(x){
   #browser()
   # need to convert the YearF from factor back to numeric value 
   pred <- cbind( x$fit@frame, pred.global=predict(x$fit, re.form=~0))
   pred$pred.stream =predict(x$fit, re.form=~(1|Stream))
   pred$Year <- as.numeric(as.character(pred$YearF))
   pred
})

# Now to generate the observed and fitted data for each Watershed/Stream combination
npages=2 # only 2 watersheds present
plyr::l_ply(1:npages, function(page){
  plot.fit <- ggplot(data=cue.data.red, aes(x=Year, y=logCUE))+
     ggtitle("Observed and fitted data from BACI in each watershed")+
     #geom_point(position=position_jitter(width=0.1), size=.1)+ too buzy
     geom_line(data=baci.pred,  aes(y=pred.global, color=CI), size=2)+
     geom_line(data=mean.CUE[ !mean.CUE$Watershed %in% watershed.exclude$Watershed,], aes(y=mean.logCUE, linetype=Stream))+
     geom_vline(data=year.restore[ !year.restore$Watershed %in% watershed.exclude$Watershed,], aes(xintercept=year.restore+.5))+
     facet_grid_paginate(Watershed+Species~AgeClass, page=page, nrow=3, ncol=3)
  plot(plot.fit)
})
```




This set of fits and estimated BACI effects can be used to construct summary plots

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
baci.effs <- plyr::ldply(baci.fits, function(x){x$BACI.diff.log})
#head(baci.effs)

#pretty up the output
temp<- baci.effs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp

ggplot(data=baci.effs, aes(y=Watershed, x=estimate))+
   ggtitle("Estimated BACI effects (log scale)",
           subtitle="BACI computed as (AT-BT)-(AC-BC)")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=lower.CL, xmax=upper.CL), height=.01)+
   xlab("Estimated BACI effect (log-scale) and 95% ci and p-value")+
   ylab("Watershed")+
   geom_vline(xintercept=0)+
   geom_text( label=format.p(baci.effs$p.value, precision=.0001), aes(x=-Inf), hjust=-.01, vjust=1.5)+
   facet_grid(Species~AgeClass)
```

A similar plot can be made for the BACI ratio (not shown). 

Case where the 95% confidence interval does not include the value of 0 (vertical line) indicate
evidence of a BACI effect. The actual p-value is also shown on the plot

## Bayesian posterior beliefs
We again sample from the posterior distribution of the parameters using the *arm* package. Some care
is needed to ensure that the BACI comparison is computed in the correct direction. Notice here that 
we must take the negative of the BACI interaction term.

```{r  }
# sample from the posterior for the BACI effect, plot the violin plots and compute the posterior belief that BACI effect >0
# Again, need to be careful that the sign of the contrast for the BACI effect is correct.
baci.post <- plyr::ldply(baci.fits, function(x){
   #browser()
   post.samples <- arm::sim(x$fit, n.sims=n.sims.post)
   data.frame(post.baci = -post.samples@fixef[,"BABefore:CIT"]) # notice that R computes before-after
})
head(baci.post)

# compute the posterior belief that difference (A-B) > 0
baci.belief <- plyr::ddply(baci.post, c("Watershed","Species","AgeClass"),
                     plyr::summarize,
                       post.belief.gt.0 = mean(post.baci >0))
head(baci.belief)

ggplot(data=baci.post, aes(y=Watershed, x=post.baci))+
   ggtitle("Posterior sample for BACI effect (log scale)",
           subtitle="BACI computed as (AT-BT)-(AC-BC)")+
   ggstance::geom_violinh()+
   geom_text(data=baci.belief, label=formatC(baci.belief$post.belief.gt.0,2, format="f"), 
             aes(x=-Inf), hjust=-.2, vjust=1.5) +
   xlab("BACI effect (log-scale) \nWith posterior belief that difference > 0")+
   ylab("Watershed ")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)
```

## Checking assumptions
It is always important to do a model assessment to ensure that the assumptions of the model are satisfied.
In this analysis the key assumptions are

* Each of the random effects (year-effects, stream-effects, the year-stream interaction effects, and site-effects) 
have a normal distribution
* There is no autocorrelation over time

It is possible to extract the estimates of the sets of random effects and create
residual and caterpillar plots.

```{r message=FALSE}
# The dianostic plot is produced for the first fit as a demonstration
plyr::l_ply(baci.fits[1], function(x){
   diag.plot <- sf.autoplot.lmer(x$fit)
   plot(diag.plot)
  
})



```

The residual plot (upper left corner) should show random scatter around 0. In this plot, 
we notice what appears to be a "line" of residuals. This is an artefact of the data because
it is not possible to have a CUE that is negative and an offset is added to all CUE values before
taking logarithms. For example, consider an observed value of CUE of 0 
where the offset was .0067 (2 fish/300 m). The observed value of $log(\mathit{CUE}+\textit{offset})=-5$.
If the predicted value on the logarithmic value was $-5$, then the residual is 0;
If the predicted value on the logarithmic scale was $-4$, then the residual is -1, etc.
As noted previously,
the distribution of the residuals at the lowest level is not that crucial because measurements at
sites are essentially being averaged for each year. 

The normal
probability plot (upper right) for the residuals (the site-effects) should fall
close to the theoretical line of fit. It is not crucial that these lowest random effects have an exact normal distribution
since they are effectively averaged in the analysis and the central-limit theorem 
indicates that the average of any distribution will tend towards normality as the
sample size increases.

The caterpillar plots should
show no large outlier effects for specific years. In this case, the estimated standard
deviation of the random effects is 0, so all of the points (with 95% credible interval)
are the same at 0.

It will be extremely difficult to detect autocorrelation over time unless you have 10+ years of 
data (Bence 1995) and so this is not done here.
```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```


# BACI analysis within each watershed on the slope
## Introduction
The BACI analysis in the previous section assumes a step change in the mean between the before
and after periods. But often the effect of restoration increases over time and so 
a BACI analysis on the slopes may be more suitable.

Here four slopes are estimates: the slope in the control streams in the *before* period; the slope in the control streams
in the *after* period using a broken stick model; the slope in the restoration stream in the *before* period, and
the slope in the restoration stream in the *after* period.

The final result from a BACI analysis on the slopes is now the **differential change** 
in the slopes of $log(CUE)$ in the usual BACI contrast.


## The statistical model
As in the before/after analysis, the analysis can be done in two ways.

* Average the multiple measurements from the sites first.
* Fit a mixed linear model to the original measurements.

Again, if the design were exactly balanced, the two approaches would give exactly the same results.

The model is similar except the $BA$ term in the BACI model of the previous section is replaced by two terms from the 
broken stick model using the
variables $Year$ and $YearP$ (as in the section on the broken stick model).

The statistical model for each watershed is
$$log(\mathit{CUE}+\textit{offset}) \sim \mathit{Year} +\mathit{YearP} + \mathit{CI} + \mathit{Year}:\mathit{CI} + \mathit{YearP}:\mathit{CI} +
\mathit{YearF}(R) + \mathit{Stream}(R)+ \mathit{YearF}:\mathit{Stream}(R)$$
where the $\mathit{Year}$ refers to the trend in the control group in the before period; 
the $\mathit{CI}$ refers to the control/impact effect;
the $\mathit{Year}:\mathit{CI}$ refers to the different slopes between control and impact in the *before* years;
the $\mathit{YearP}$ is the change in slope for the control group in the *after* period relative to the *before* period;
and the $\mathit{YearP}:CI$ is the BACI effect (the differential change in the slopes) which is the term of interest;
the $\mathit{YearF}$ refers to the year random effects (it must be declared as a factor to avoid being 
confused with a linear trend);
the $\mathit{Stream}$ term refers to the (random) differences among streams;
and the $\mathit{YearF}:\mathit{Stream}$ refers to the random effect where streams do not 
track each other in parallel over time.

The effects of stream will cancel in the analysis because each stream is measured
before and after impact, i.e., each stream serves as a blocking variable.
The effects of year will also cancel in the analysis because in each year, both control
and impact streams are measured, i.e., each year also serves as a blocking variable.
The last term represents inconsistency in the mean of a stream over time compared to other streams, i.e.,
a non-parallelism effect and is often the limiting factor in a power analysis 
in a BACI design.

The same model can be used if some streams are not measured in some years and if the design were unbalanced.


## Using R to fit and extract results
We exclude the watersheds that don't have a control and treatment stream and a replicate of either the control
or treatment streams.

```{r }
n.streams <- plyr::ddply(cue.data, "Watershed", plyr::summarize,
                         n.control  =length(unique(Stream[ CI=="C"])),
                         n.treatment=length(unique(Stream[ CI=="T"])))
n.streams

watershed.exclude <- n.streams[ n.streams$n.control==0 | n.streams$n.treatment==0,]
watershed.exclude

dim(cue.data)
cue.data.red <- cue.data[ !cue.data$Watershed %in% watershed.exclude$Watershed, ]
dim(cue.data.red)
```

The $\mathit{lmerTest}$ package is used to fit the model; the $\mathit{emmeans}$ package is used to estimate the
BACI effect on the slopes after the model is fit. In this case, the $\mathit{emtrends}$ function provides estimates
of the slopes and changes in the slopes for the control and restoration streams. Because the coefficient of the
$YearP$ term is already the change in slopes between the $after$ and $before$ periods, we only need the pairwise
difference of these coefficients to estimate the BACI effect on the slopes. This corresponds to the
BACI contrast 
$$\mathit{BACI} = (\beta_{\mathit{After~C}} - \beta_{\mathit{Before~C}}) - (\beta_{\mathit{After~T}} - \beta_{\mathit{Before~T}})$$
where the $\beta$ terms refer to the SLOPES in the combination of BA and CI.

```{r  warning=FALSE}
# Get all of the fits and store them in a list with the emmeans computatons
baci.slope.fits <- plyr::dlply(cue.data.red, c("Watershed","Species","AgeClass"), function(x){
   #browser()
   x$YearF <- as.factor(x$Year)
   year.restore <- min(x$Year[x$BA=="After"])
   x$YearP = pmax(0, x$Year-year.restore)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ Year+YearP + CI + Year:CI + YearP:CI + (1|YearF) +(1|Stream) +(1|YearF:Stream), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emtrends(fit, "CI", var="YearP") # This are the two changes in the slopes for C and I
   BACI.diff.log <- summary(pairs(fit.emmo), infer=TRUE)
   BACI.ratio    <- NA  # this doesn't make sense in this context
   list(fit=fit, 
        BACI.diff.log=BACI.diff.log, 
        BACI.ratio=BACI.ratio)
})

# Show the results for the first fit
baci.slope.fits[[1]]$fit
baci.slope.fits[[1]]$BACI.diff.log
```
A message about a singular fit indicates that the estimated standard deviation of some of the random effects is 0.
The estimated BACI effect on the slopes (i.e. the differential change in the slopes)
along with standard errors, a 95% confidence limit,
and p-value of the test for no effect is produced.

We can examine the plots of fitted and predicted values to check the general form of the fit and
again for any obvious problems:
```{r}
# Get the fitted values
baci.slope.pred <- plyr::ldply(baci.slope.fits, function(x){
   #browser()
   # need to convert the YearF from factor back to numeric value 
   pred <- cbind( x$fit@frame, pred.global=predict(x$fit, re.form=~0))
   pred$pred.stream =predict(x$fit, re.form=~(1|Stream))
   pred$Year <- as.numeric(as.character(pred$YearF))
   pred
})

# Now to generate the observed and fitted data for each Watershed/Stream combination
npages=2 # only 2 watersheds present
plyr::l_ply(1:npages, function(page){
  plot.fit <- ggplot(data=cue.data.red, aes(x=Year, y=logCUE))+
     ggtitle("Observed and fitted data from BACI on slopes in each watershed")+
     #geom_point(position=position_jitter(width=0.1), size=.1)+ too buzy
     geom_line(data=baci.slope.pred,  aes(y=pred.global, color=CI), size=2)+
     geom_line(data=mean.CUE[ !mean.CUE$Watershed %in% watershed.exclude$Watershed,], aes(y=mean.logCUE, linetype=Stream))+
     geom_vline(data=year.restore[ !year.restore$Watershed %in% watershed.exclude$Watershed,], aes(xintercept=year.restore+.5))+
     facet_grid_paginate(Watershed+Species~AgeClass, page=page, nrow=3, ncol=3)
  plot(plot.fit)
})
```




This set of fits and estimated BACI effects can be used to construct summary plots

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
baci.slope.effs <- plyr::ldply(baci.slope.fits, function(x){x$BACI.diff.log})
#head(baci.effs)

#pretty up the output
temp<- baci.slope.effs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp

ggplot(data=baci.slope.effs, aes(y=Watershed, x=estimate))+
   ggtitle("Estimated BACI effects on slope (log scale)",
           subtitle="BACI computed as (AC-BC)-(AT-BT)")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=lower.CL, xmax=upper.CL), height=.01)+
   xlab("Estimated BACI effect (log-scale) and 95% ci and p-value")+
   ylab("Watershed")+
   geom_vline(xintercept=0)+
   geom_text( label=format.p(baci.slope.effs$p.value, precision=.001), aes(x=-Inf), hjust=-.01, vjust=1.5)+
   facet_grid(Species~AgeClass)
```


Case where the 95% confidence interval does not include the value of 0 (vertical line) indicate
evidence of a BACI effect. The actual p-value is also shown on the plot

## Bayesian posterior beliefs
We again sample from the posterior distribution of the parameters using the *arm* package. Some care
is needed to ensure that the BACI comparison is computed in the correct direction. Notice here that 
we must take the negative of the BACI interaction term.

```{r  }
# sample from the posterior for the BACI effect, plot the violin plots and compute the posterior belief that BACI effect >0
# Again, need to be careful that the sign of the contrast for the BACI effect on slopes is correct.
baci.slope.post <- plyr::ldply(baci.slope.fits, function(x){
   #browser()
   post.samples <- arm::sim(x$fit, n.sims=n.sims.post)
   data.frame(post.baci = -post.samples@fixef[,"YearP:CIT"]) # notice that R computes before-after
})
head(baci.slope.post)

# compute the posterior belief that difference (A-B) > 0
baci.slope.belief <- plyr::ddply(baci.slope.post, c("Watershed","Species","AgeClass"),
                     plyr::summarize,
                       post.belief.gt.0 = mean(post.baci >0))
head(baci.slope.belief)

ggplot(data=baci.slope.post, aes(y=Watershed, x=post.baci))+
   ggtitle("Posterior sample for BACI effect on slopes (log scale)",
           subtitle="BACI computed as (AC-BC)-(AT-BT)")+
   ggstance::geom_violinh()+
   geom_text(data=baci.slope.belief, label=formatC(baci.slope.belief$post.belief.gt.0,2, format="f"), 
             aes(x=-Inf), hjust=-.2, vjust=1.5) +
   xlab("BACI effect on slopes (log-scale) \nWith posterior belief that difference > 0")+
   ylab("Watershed ")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)
```

## Checking assumptions
It is always important to do a model assessment to ensure that the assumptions of the model are satisfied.
In this analysis the key assumptions are

* Each of the random effects (year-effects, stream-effects, the year-stream interaction effects, and site-effects) 
have a normal distribution
* There is no autocorrelation over time

It is possible to extract the estimates of the sets of random effects and create
residual and caterpillar plots.

```{r message=FALSE}
# The dianostic plot is produced for the first fit as a demonstration
plyr::l_ply(baci.slope.fits[1], function(x){
   diag.plot <- sf.autoplot.lmer(x$fit)
   plot(diag.plot)
  
})



```

The residual plot (upper left corner) should show random scatter around 0. In this plot, 
we notice what appears to be a "line" of residuals. This is an artefact of the data as noted in the
previous section. 

The normal
probability plot (upper right) for the residuals (the site-effects) should fall
close to the theoretical line of fit. It is not crucial that these lowest random effects have an exact normal distribution
since they are effectively averaged in the analysis and the central-limit theorem 
indicates that the average of any distribution will tend towards normality as the
sample size increases.

The caterpillar plots should
show no large outlier effects for specific years. In this case, the estimated standard
deviation of the random effects is 0, so all of the points (with 95% credible interval)
are the same at 0.

It will be extremely difficult to detect autocorrelation over time unless you have 10+ years of 
data (Bence 1995) and so this is not done here.




```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```
# Multiple BACI - combining over watersheds
## Introduction
The impact of restoration may not be equal for all watersheds, i.e. variation may also exist in the BACI effect.
In this section, we will fit a super-model for all streams that estimates the average effect of restoration
across watersheds, but also how variable this effect is across watersheds.


## The statistical model
We once again use a mixed linear model to account for the various sources of variation in the model.

The statistical model for each watershed is
$$log(\mathit{CUE}+\textit{offset}) \sim \mathit{BA} + \mathit{CI} + \mathit{BA}:\mathit{CI} + 
\mathit{Watershed}:\mathit{CI}(R) +\mathit{Watershed}:\mathit{BA}(R)+ \mathit{Watershed}:\mathit{CI}:\mathit{BA})+
\mathit{YearF}(R) + \mathit{Stream}(R)+ \mathit{YearF}:\mathit{Stream}(R)$$
where the new random effect involving the *Watershed* allow for variation in these effects
across the watershed.

It is not necessary to have a random effect for Watershed because we have random effects
for each individual stream.

It is not necessary that the restoration year is the same in all watershed as long as the
data file indicates which year belong to which period for each watershed.

We implicitly assume that year-specific effects apply similarly to all watersheds. However, watershed differences in the year-effect can be 
assessed by adding a watershed-year interaction and looking at the size of the resulting
variance component.


## Using R to fit and extract results
It would make no sense to combine results over watersheds where only one watershed is involved
for a species-age class. For example, *RNTR* is only observed in one watershed.

Similarly, we need to exclude watersheds where no BACI design was done. 

```{r }
# exclude species that only appear in 1 watershed
n.watersheds <- plyr::ddply(cue.data, c("Species"), plyr::summarize,
                         n.watersheds=length(unique(Watershed)))
n.watersheds

species.exclude <- n.watersheds[ n.watersheds$n.watersheds==1,]
species.exclude

dim(cue.data)
cue.data.red <- cue.data[ !cue.data$Species %in% species.exclude$Species, ]
dim(cue.data.red)

# exclude watersheds that don't have a BACI design
n.streams <- plyr::ddply(cue.data.red, "Watershed", plyr::summarize,
                         n.control  =length(unique(Stream[ CI=="C"])),
                         n.treatment=length(unique(Stream[ CI=="T"])))
n.streams

watershed.exclude <- n.streams[ n.streams$n.control==0 | n.streams$n.treatment==0,]
watershed.exclude

dim(cue.data.red)
cue.data.red <- cue.data.red[ !cue.data.red$Watershed %in% watershed.exclude$Watershed, ]
dim(cue.data.red)
```

The $\mathit{lmerTest}$ package is used to fit the model; the $\mathit{emmeans}$ package is used to estimate the
BACI effect after the model is fit. Some care is needed to specify the coefficients of the BACI contrast 
properly. For example, in this model, the interaction (BACI) term is specified as $\mathit{BA}:\mathit{CI}$.
The $\mathit{BA}$ variable is coded using *After* and *Before*; the $\mathit{CI}$ variable is coded using *C*
and *I*. Consequently, the BACI terms are in the order *After C*, *Before C*,*After T*, and *Before T*.

We would like the BACI contrast to be computed as the difference of difference in the mean as
$$\mathit{BACI} = (\mu_{\mathit{After~T}} - \mu_{\mathit{Before~T}}) - (\mu_{\mathit{After~C}} - \mu_{\mathit{Before~C}})$$
and so the BACI coefficients are $(-1,1,1,-1)$ as specified below.

```{r  warning=FALSE}
# Get all of the fits for the global BACI model and store them in a list with the emmeans computatons
baci.g.fits <- plyr::dlply(cue.data.red, c("Species","AgeClass"), function(x){
   #browser()
   x$YearF <- factor(x$Year)
   # If we use the log() in the fit, the emmeans package can do the back transform automatically for us
   fit <- lmerTest::lmer(log(CUE+offset) ~ BA + CI + BA:CI + 
                           (1|Watershed:CI)+(1|Watershed:BA)+(1|Watershed:CI:BA)+
                           (1|YearF) +(1|Stream) +(1|YearF:Stream), data=x)
   # we create the emmeans object and then find the pairwise difference
   fit.emmo <- emmeans::emmeans(fit, ~BA:CI)
   BACI.diff.log <- summary(contrast(fit.emmo,list(baci=c(-1,1,1,-1))), infer=TRUE)
   BACI.ratio    <- summary(contrast(fit.emmo,list(baci=c(-1,1,1,-1)), type="response")) 
   list(fit=fit, 
        BACI.diff.log=BACI.diff.log, 
        BACI.ratio=BACI.ratio)
})

# Show the results for one of the fits
baci.fits[[1]]$fit
baci.fits[[1]]$BACI.diff.log
baci.fits[[1]]$BACI.ratio
```
A message about a singular fit indicates that the estimated standard deviation of some of the random effects is 0.
The estimated BACI effect along with standard errors, a 95% confidence limit,
and p-value of the test for no effect is produced.

We can examine the plots of fitted and predicted values to check the general form of the fit and
again for any obvious problems:
```{r}
# Get the fitted values
baci.g.pred <- plyr::ldply(baci.g.fits, function(x){
   #browser()
   # need to convert the YearF from factor back to numeric value 
   pred <- cbind( x$fit@frame, pred.global=predict(x$fit, re.form=~0))
   pred$pred.watershed =predict(x$fit, re.form=~(1|Watershed:CI))

   pred$pred.stream =predict(x$fit, re.form=~(1|Stream))
   pred$Year <- as.numeric(as.character(pred$YearF))
   pred
})

# Now to generate the observed and fitted data for each Watershed/Stream combination
npages=1 # only 1 page is needed
plyr::l_ply(1:npages, function(page){
  plot.fit <- ggplot(data=cue.data.red, aes(x=Year, y=logCUE))+
     ggtitle("Observed and fitted data from global BACI")+
     #geom_point(position=position_jitter(width=0.1), size=.1)+ too buzy
     geom_line(data=baci.g.pred,  aes(y=pred.global, color=CI), size=2)+
     geom_line(data=mean.CUE[ !mean.CUE$Watershed %in% watershed.exclude$Watershed,], aes(y=mean.logCUE, linetype=Stream))+
     geom_vline(data=year.restore[ !year.restore$Watershed %in% watershed.exclude$Watershed,], aes(xintercept=year.restore+.5))+
     facet_grid_paginate(Species~AgeClass, page=page, nrow=3, ncol=3)
  plot(plot.fit)
})
```

This set of fits and estimated BACI effects can be used to construct summary plots

```{r  }
# Get all of the fits and store them in a list with the emmeans computatons
baci.g.effs <- plyr::ldply(baci.g.fits, function(x){x$BACI.diff.log})
#head(baci.effs)

#pretty up the output
temp<- baci.g.effs
num.col <- sapply(temp, is.numeric)
temp[,num.col] <- round(temp[,num.col],2)
temp$t.ratio <- NULL
temp

ggplot(data=baci.g.effs, aes(y="GLOBAL", x=estimate))+
   ggtitle("Estimated GLOBAL BACI effects (log scale)",
           subtitle="BACI computed as (AT-BT)-(AC-BC)")+
   geom_point()+
   ggplot2::geom_errorbarh( aes(xmin=lower.CL, xmax=upper.CL), height=.01)+
   xlab("Estimated BACI effect (log-scale) and 95% ci and p-value")+
   ylab("Watershed")+
   geom_vline(xintercept=0)+
   geom_text( label=format.p(baci.g.effs$p.value, precision=.001), aes(x=-Inf), hjust=-.01, vjust=1.5)+
   facet_grid(Species~AgeClass)
```

A similar plot can be made for the BACI ratio (not shown). 

Case where the 95% confidence interval does not include the value of 0 (vertical line) indicate
evidence of a GLOBAL BACI effect. The actual p-value is also shown on the plot.

The BACI effect may vary among the watersheds. This is measured by the standard deviation of the
BACI by watershed interaction:
```{r }
# Extract the BACI x watershed standard deviation
baci.g.w.baci.sd <- plyr::ldply(baci.g.fits, function(x){
    #browser()
    vc <- as.data.frame(VarCorr(x$fit))
    data.frame(w.baci.sd = vc[ vc$grp=="Watershed:CI:BA", "sdcor"])
})
baci.g.w.baci.sd
```

This provides information on the variation in the BACI effect among the watersheds.

## Bayesian posterior beliefs
We again sample from the posterior distribution of the parameters using the *arm* package. Some care
is needed to ensure that the BACI comparison is computed in the correct direction. Notice here that 
we must take the negative of the BACI interaction term.

```{r  }
# sample from the posterior for the BACI effect, plot the violin plots and compute the posterior belief that BACI effect >0
# Again, need to be careful that the sign of the contrast for the BACI effect is correct.
baci.g.post <- plyr::ldply(baci.g.fits, function(x){
   #browser()
   post.samples <- arm::sim(x$fit, n.sims=n.sims.post)
   data.frame(post.baci = -post.samples@fixef[,"BABefore:CIT"]) # notice that R computes before-after
})
head(baci.g.post)

# compute the posterior belief that difference (A-B) > 0
baci.g.belief <- plyr::ddply(baci.g.post, c("Species","AgeClass"),
                     plyr::summarize,
                       post.belief.gt.0 = mean(post.baci >0))
head(baci.g.belief)

ggplot(data=baci.g.post, aes(y="GLOBAL", x=post.baci))+
   ggtitle("Posterior sample for GLOBAL BACI effect (log scale)",
           subtitle="BACI computed as (AT-BT)-(AC-BC)")+
   ggstance::geom_violinh()+
   geom_text(data=baci.g.belief, label=formatC(baci.g.belief$post.belief.gt.0,2, format="f"), 
             aes(x=-Inf), hjust=-.2, vjust=1.5) +
   xlab("BACI effect (log-scale) \nWith posterior belief that difference > 0")+
   ylab("Watershed ")+
   geom_vline(xintercept=0)+
   facet_grid(Species~AgeClass)
```



## Checking assumptions

It is possible to extract the estimates of the sets of random effects and create
residual and caterpillar plots in the usual way.

```{r message=FALSE}
# The dianostic plot is produced for the one of the fits as a demonstration
plyr::l_ply(baci.g.fits[1], function(x){
   diag.plot <- sf.autoplot.lmer(x$fit)
   plot(diag.plot)
  
})
```

The residual plots are interpreted in the same way as previously.
Again, the "line" in the residuals is an artefact of CUE being limited below by 0 and the offset. As noted previously,
the distribution of the residuals at the lowest level is not that crucial because measurements at
sites are essentially being averaged for each year. 





```{r echo=FALSE, eval=FALSE}
####################################################################################################################
####################################################################################################################
####################################################################################################################
####################################################################################################################
```

# Summary and recommendations
A key point when analyzing the future data is to account for stream-years-sites where no fish of a particular
species was found, i.e. 0 counts must be imputed. This is a common problem with many database systems
that only record positive counts. An additional table is needed in the database of the set of sites visited in
each stream-year and the species to be expected, so that 0's can be properly imputed. The FWMIS system has been
programmed to do the 0 imputations, so no further imputation will be needed.

This demonstration shows several ways to analyse the upcoming data to be collected from the restoration
program. The analysis become progressively more complex so deficiencies within each step are addressed.

* The before-after analysis estimate of effect size for each stream will be confounded with the effects
of any other temporal variable. For example, suppose that in the five years after restoration takes place,
that the precipitation levels are much lower than normal leading to degradation of water quality and
 a reduction in CUE. This will be confounded with the difference in mean response between the after
 and before periods.
 
 * The before-after analysis also assumes that the change due to restoration is an immediate step function.
 In some cases, improvements after restoration will be gradual and cumulative as, for example, 
 habitat slowly improves. In this case, a broken-stick model looks for a change in the trend
 starting at the time of restoration. In this case, the time of the change in trend is known. 
 It is a much more difficult problem to fit a model where the break-point is to be estimated as well.
 This is known as the change-point problem (Toms and Lesperance, 2003) and will require many, many, many
 more years of data before it is feasible (typically at least 30-40 years of data!).
 
 * The BACI design uses stream(s) that have not been restored as controls to account for temporal change.
 In this study, multiple control streams provide assurances that the observed effects are not 
 specific to a particular control stream but can be generalized across the set of control streams.
 Only having a single treatment stream, unfortunately, still leaves some question that the results
 are particular to this specific treatment stream, but the variability among control streams provides
 some (limited) information on the variation that would be expected if multiple treatment streams
 were used.
 
 * Finally, a global BACI analysis combined results over watersheds. Here the BACI effect is a 
 "global" average effect. Because the restoration actions are likely quite different across
 watersheds, this average BACI effect must be interpreted with care.
 
This demonstration shows the results from a classical analysis using linear mixed models fit with REML,
the standard analysis that can account for multiple sources of variation. In particular, any analysis 
that involves measurements over time must account for year-specific effects (process error) which, if
ignored, would violate the assumption of independence among observations. In this particular case,
sites within a stream-year are pseudo-replicates and cannot be analyzed directly.

A common problem with the results from Neyman-Pearson approaches (p-values and confidence intervals), is that
these statistics do not have an intuitive interpretation. For example, many people will interpret a p-value of 0.01 as
being the probability that the hypothesis of no effect is true. This is an erroneous interpretation. 
Similarly, many people interpret confidence intervals as a zone of belief of the underlying parameter value.
This is again erroneous. In both cases, a Bayesian perspective is helpful, in particular, the samples from 
the posterior distribution of a parameter can be interpreted as the "belief" in particular values of the 
parameters. So it is sensible to compute the posterior belief that the effect is positive.
The posterior belief can be interpreted as a measure of the weight of evidence in favor of the hypothesis.
Fortunately, 
the *arm* package in $R$ makes it particularly easy to sample from the posterior distribution of 
parameters in mixed linear models and so easy to compute these beliefs.

# References
AEP (Alberta Environment and Parks). 2018. Electrofishing standard for sampling of rivers in Alberta. Fisheries Management, Policy and Operations Branch, AEP. April 2018. 24 p.

Bence, J. R. 1995. Analysis of Short Time Series: Correction for Autocorrelation. Ecology 76: 62839.

Kuznetsova A, Brockhoff PB, Christensen RHB (2017). lmerTest Package: Tests in Linear Mixed Effects Models. Journal
of Statistical Software, 82, 1-26. doi: 10.18637/jss.v082.i13 (URL: http://doi.org/10.18637/jss.v082.i13).

Gelman, A. and Hill, J. (2006). Data Analysis Using Regression and Multilevel/Hierarchical Models. Cambridge University Press.

Gelman, A. and Su, Yu-Sung (2018). arm: Data Analysis Using Regression and Multilevel/Hierarchical Models. R package
  version 1.10-1. https://CRAN.R-project.org/package=arm

Lenth, R. (2019). emmeans: Estimated Marginal Means, aka Least-Squares Means. R package version 1.3.2.
  https://CRAN.R-project.org/package=emmeans
  
R Core Team (2019). R: A language and environment for statistical computing. R Foundation for Statistical Computing,
  Vienna, Austria. URL https://www.R-project.org/.
  
Schwarz. C. J. (2018). Power analysis for evaluating the effectiveness of intensive OHV reclamation activities to recover native trout along the East Slopes of Alberta, Canada. Prepared for the Alberta Ministry of Environment. 53pp. 

Schwarz, C. J. (2017). Monitoring Design and Analysis using the Fish Sustainability Index (FSI). Prepared for the Alberta Ministry of Environment. 16 pp + Extensive R code.

Toms, J. D., and M. L. Lesperance. 2003. Piecewise Regression A Tool for Identifying Ecological Thresholds. Ecology 84 (8): 203441.
